---
title: "Unequal Exposure: Redlining’s Legacy on Los Angeles’ Environmental Health and Biodiversity"
author: "Karol Paya"
date: Dec 7, 2024
execute: 
  eval: True
  warning: False
  message: False
format:
  html:
    toc: true
editor_options: 
  chunk_output_type: console
---

# Study Background

Redlining is a discriminatory practice that continues to shape the environmental and social landscapes of many U.S. cities. It involves the Home Owners’ Loan Corporation (HOLC) ranking neighborhoods based on perceived investment risk, with red areas being deemed hazardous and systematically denied access to services. This practice leads to significant disparities in environmental quality, with redlined neighborhoods often facing poorer living conditions, less green space, and higher exposure to pollution [(New York Times, 2021)](https://www.nytimes.com/2021/08/17/realestate/what-is-redlining.html). Recent research, including a study by Ellis-Soto et al. (2023), has highlighted how this historical injustice extends into the realm of biodiversity. Specifically, redlined neighborhoods are disproportionately undersampled in citizen science data, which is used to inform conservation decisions. This gap is concerning because it could result in biodiversity conservation strategies that overlook or inadequately address the needs of communities historically affected by redlining.

This study assesses whether these findings hold true for Los Angeles by analyzing the distribution of bird observations from 2022, comparing the volume of data collected in redlined versus non-redlined areas, and exploring the implications for environmental justice.

# Data Details

EJScreen is a tool provided by the U.S. Environmental Protection Agency that offers detailed environmental and demographic data for the United States, available at both the Census tract and block group levels. The dataset used in this study specifically contains data at the block group level, providing a granular view of environmental justice indicators across communities.

The Mapping Inequality project, led by the Digital Scholarship Lab at the University of Richmond, has digitized historical maps and data from the Home Owners' Loan Corporation (HOLC) redlining surveys.

The Global Biodiversity Information Facility (GBIF) is the largest aggregator of biodiversity observations worldwide. The dataset includes information on species sightings, typically recording the location and date of each observation.

## References:

\*EJScreen Dataset\*

United States Environmental Protection Agency. (2024). EJScreen: Environmental Justice Screening and Mapping Tool. Retrieved October 13, 2024, from https://www.epa.gov/ejscreen/download-ejscreen-data

\*HOLC Redlining\*

University of Richmond - Digital Scholarship Lab. (n.d.). Mapping Inequality: Redlining in New Deal America. Retrieved October 13, 2024, from https://dsl.richmond.edu/

\*Biodiversity Observations\*

Global Biodiversity Information Facility. (n.d.). GBIF Dataset Search. Retrieved October 13, 2024, from https://www.gbif.org/dataset/search?q=

# Geospatial Assessment: Impact on Environmental Justice and Biodiversity in Los Angeles

## Load Libraries

<details>

<summary>Click to view code</summary>

```{r}
# Load libraries
library(sf)
library(tmap)
library(here)
library(tidyverse)
library(dplyr)
library(gt)
```

</details>

## Load Data

<details>

<summary>Click to view code</summary>

```{r}
# Loading data
ejscreen <- sf::read_sf(here::here("data", "ejscreen",
                                   "EJSCREEN_2023_BG_StatePct_with_AS_CNMI_GU_VI.gdb"))

holc <- sf::st_read(here::here("data", "mapping-inequality", 
                                 "mapping-inequality-los-angeles.json"))

birds <- sf::st_read(here::here("data","gbif-birds-LA","gbif-birds-LA.shp"))
```

</details>

## Check Coordinate System Match

<details>

<summary>Click to view code</summary>

```{r}
# Check holc map and ejscreen have the same coordinate systems
if(st_crs(holc) == st_crs(ejscreen)){
  print("it's a match")
} else {
  print("Not a match, reprojecting..")
  holc_rpr <- st_transform(holc, st_crs(ejscreen)) # Adjust coordinate system
}
```

```{r}
# Check if birds & ejscreen coordinate systems are the same
if(st_crs(birds) == st_crs(ejscreen)){
  print("it's a match")
} else {
  print("Not a match, reprojecting..")
  birds_rpr <- st_transform(birds, st_crs(ejscreen)) # Adjust coordinate system
}
```

</details>

## Map of Historical Redlining Neighborhoods in LA County

The map below of Los Angeles County, visualizes the historical Home Owners' Loan Corporation (HOLC) grades assigned to different neighborhoods. It uses the tmap package to render the map, setting the mode to interactive viewing to allow for zooming and panning.

<details>

<summary>Click to view code</summary>

```{r}
# Create a map of LA county showing the HOLC Grades
map1<- tm_shape(holc_rpr) +
  tm_polygons("grade", #Assign colors to each grade
               palette = c("A" = "chartreuse3", 
                           "B" = "blue", 
                           "C" = "yellow", 
                           "D" = "red"),
               title = "HOLC Grade") +
  tm_layout(title = "Map of Los Angeles County by Grade") # Add title
```

</details>

```{r}
# Plot HOLC map of LA
tmap_mode("view") # tmap mode set to interactive viewing
tmap_options(check.and.fix = TRUE)

map1
```

The following section of the analysis filters the ejscreen dataset to focus on Los Angeles County, selecting relevant variables such as low-income percentage, particulate matter levels, and life expectancy percentiles. A new column is created to extract the block group code from the ID, which is then used to perform a spatial intersection with the HOLC dataset (holc_rpr). The resulting LA_holc dataset is grouped by HOLC grade and census block group, and the percentage of census blocks within each HOLC grade is calculated. A summary table is created, presenting the HOLC grade, block group code, and the corresponding percentage of blocks within each grade.

<details>

<summary>Click to view code</summary>

```{r}
# Filter LA County
LA_ejscreen <- ejscreen  %>%
  filter(CNTY_NAME == "Los Angeles County") %>%
  select(ID, LOWINCPCT, P_PM25, P_LIFEEXPPCT)%>%
  mutate(Block_Group_Code = substr(ID, nchar(ID), nchar(ID)))

# Perform spatial joint of LA_ejscreen & HOLC dataset
LA_holc <- st_intersection(LA_ejscreen, holc_rpr)

# Create summary table of the % census block per HOLC grade
block_holc<- LA_holc %>%
  group_by(grade,Block_Group_Code) %>%
  summarize(count = n()) %>%
  mutate(percent = count / sum(count) * 100)

# Create new dataframe without shape attributes
censusblock_grade_table <- data.frame(
  Grade = block_holc$grade,
  Block_group = block_holc$Block_Group_Code,
  Percent = block_holc$percent
)

# Use library gt to create a table to better present the results

gt_table <- censusblock_grade_table %>%
  gt() %>%
  tab_header(
    title = "Census Block Group Percentages by HOLC Grade"
  ) %>%
  cols_label(
    Grade = "HOLC Grade",
    Block_group = "Census Block Group",
    Percent = "Percent (%)"
  ) %>%
  fmt_number(
    columns = c(Percent),
    decimals = 2
)
```

</details>

```{r}
# Display the table
gt_table
```

This summary table displays the average values of various conditions (low-income percentage, particulate matter levels, and life expectancy percentage) for each HOLC grade. The LA_holc dataset is grouped by HOLC grade, and the mean values for each condition are calculated, excluding any missing values. The data is then transformed into a long format using pivot_longer, making it easier to present the different conditions.

<details>

<summary>Click to view code</summary>

```{r}
# Create summary table of the calculated averages of different conditions
block_holc_conditions<- LA_holc %>%
  group_by(grade)%>%
  summarize(
    mean_lowincome_percent = mean(LOWINCPCT, na.rm = TRUE),
    mean_PM25_pct = mean(P_PM25, na.rm = TRUE),
    mean_lifeexpectancy_pct = mean(P_LIFEEXPPCT, na.rm = TRUE)
  )

# Convert block_holc_conditions to a long format
block_holc_conditions_long<- block_holc_conditions %>%
  pivot_longer(cols = starts_with("mean"), 
               names_to = "conditions", 
               values_to = "averages"
)

# Create new dataframe without shape attributes
block_hold_table <- data.frame(
  Grade = block_holc_conditions_long$grade,
  Conditions = block_holc_conditions_long$conditions,
  Averages = block_holc_conditions_long$averages
)

# Use library gt to create a table to better present the results
gt_table2 <- block_hold_table %>%
  gt() %>%
  tab_header(
    title = "HOLC Grade Conditions Summary"
  ) %>%
  cols_label(
    Grade = "HOLC Grade",
    Conditions = "Conditions",
    Averages = "Average"
  ) %>%
  fmt_number(
    columns = c(Averages),
    decimals = 2
)
```

</details>

```{r}
# Display the gt table
gt_table2
```

The "Condition Averages by HOLC Grades" bar plot displays the average values of different conditions (low income percentage, life expectancy percentiles, and particulate matter levels) across neighborhoods classified by their HOLC grade. This visualization helps to highlight disparities in these conditions between different areas, potentially revealing the ongoing impact of redlining.

<details>

<summary>Click to view code</summary>

```{r}
# Create bar plot of condition averages across different grades
bar_plot<-ggplot(block_holc_conditions_long, 
       aes(x = conditions, 
           y = averages, 
           fill = grade)) +
  geom_bar(stat = "identity", 
           position = "dodge") +
  labs(title="Condition Averages by HOLC Grades",
       x = "Condition", 
       y = "Averages", fill = "HOLC Grade") +
  scale_x_discrete(labels = c("Low Income %",
                              "Life Expectancy \nPercentiles (Years)",
                              "Particulate Matter 2.5 \n Percentiles (µg/m³)")) +
  scale_fill_manual(values = c("A" = "chartreuse3", 
                                "B" = "blue", 
                                "C" = "yellow", 
                                "D" = "red"))+
  theme_minimal()
```

</details>

```{r}
# Display bar plot
bar_plot
```

When comparing the mean condition percentiles across each grade, no clear and logical relationship was identified between life expectancy and PM2.5 levels. The data hows that higher PM2.5 concentrations correlated with increased life expectancy across the different grades.

When examining the low-income percentiles and their corresponding grades, the results indicate a consistent trend: Grade A has the lowest average population living in low income, with this figure increasing as we move through the other HOLC grading systems.

## Legacy of Redlining in Biodiversity Observations

This section provides a summary of the percentage of biodiversity observations within redlined neighborhoods, categorized by their corresponding HOLC grade. To accurately analyze the spatial overlap between the biodiversity observations and the historical redlining boundaries, the initial use of `st_join` was found to result in additional blank rows due to non-overlapping datasets. As a solution, the `st_intersection` function will be used.

```{r}
# Compute st_intersection and check number of rows
bird_holc <- st_intersection(holc_rpr, birds_rpr)
print(paste("Number of Rows for st_intersection:", nrow(bird_holc)))

# Compute a st_join and check number of rows
bird_holc2<- st_join(holc_rpr, birds_rpr,join=st_intersects)
print(paste("Number of Rows for st_join:", nrow(bird_holc2)))
```

</details>

This section generates a summary of bird observations within neighborhoods categorized by their historical HOLC grades. The bird_holc dataset is grouped by grade, and the total count of bird observations per grade is calculated, followed by the computation of the percentage of total observations for each grade. This part of the analysis provides insight into how biodiversity observations are distributed across neighborhoods with different historical designations, which could offer clues into the lasting effects of redlining on biodiversity monitoring.

<details>

<summary>Click to view code</summary>

```{r}
# Create summary table of the percent bird observation per HOLC grade
bird_obs<- bird_holc %>%
  group_by(grade) %>%
  summarize(count = n()) %>%
  mutate(percent_observations = count / sum(count) * 100)

# Create new dataframe without shape attributes
bird_table <- data.frame(
  Grade = bird_obs$grade,
  Count = bird_obs$count,
  percent_observation = bird_obs$percent_observations
)

# Create gt table for bird observation percentages
bird_table%>%
  gt() %>%
  tab_header(
    title = "HOLC Grade Conditions Summary"
  ) %>%
  cols_label(
    Grade = "HOLC Grade",
    Count = "Bird Observation Count",
    percent_observation = "Percent Observation (%)"
  ) %>%
  fmt_number(
    columns = c(percent_observation),
    decimals = 2
)

# Create Bar graph showing the percentages of bird observation per HOLC grade
bar_plot2<- ggplot(bird_obs, aes(x = grade, y = percent_observations, fill = grade)) +
  geom_bar(stat = "identity",show.legend = FALSE) +
  labs(title = "Bird Observations Percentage by HOLC Grade",
       x = "HOLC Grade",
       y = "Percent (%)") +
  scale_fill_manual(values = c("A" = "chartreuse3", 
                           "B" = "blue", 
                           "C" = "yellow", 
                           "D" = "red")) +
  theme_minimal()
```

</details>

```{r}
# Display bar plot
bar_plot2
```

## Study Findings

The results of the analysis do not fully align with the findings of Ellis-Soto et al. (2023). While their study suggested that redlined neighborhoods, particularly those with lower HOLC grades, are typically undersampled in terms of biodiversity data, our analysis shows that grade C neighborhoods had the highest number of bird observations. This discrepancy could reflect biases in data collection, or it might indicate the influence of ecological factors that warrant further investigation.

The findings suggest that a direct relationship between bird observations and HOLC grades may not be as clear-cut as previously thought. Therefore, additional research, including a more comprehensive historical dataset on biodiversity in redlined areas, is necessary to better understand the impact of redlining on species observations and to assess the validity of Ellis-Soto et al.'s conclusions.
